Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population

### 2024-04-16 analysis/paper/paper.tex
add tex file for submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=10mm,bottom=10mm,left=30mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{fancyhdr, graphicx}
\usepackage[hidelinks]{hyperref}

\fancyhead{\vspace{70pt}}
\renewcommand{\headrulewidth}{1.8pt}

\makeatletter
\def\@maketitle{%
  \newpage
  \hfill\href{https://doi.org/10.24072/pci.archaeo.100389}{\includegraphics[width=3.5cm]{badge_PCI_Archaeology.png}}\hfill
  \null
 \begin{center}
{\LARGE
\textbf\newline{\@title}
}
%\newline

\end{center}
  \par}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Multiproxy analysis exploring patterns of diet and disease in
dental calculus and skeletal remains from a 19th century Dutch
population}
\author{Bjørn Peare Bartholdy \and Jørgen B. Hasselstrøm \and Lambert K.
Sørensen \and Maia Casna \and Menno Hoogland \and Historisch Genootschap
Beemster \and Amanda G. Henry}
\date{}

\begin{document}
%% modified from: https://github.com/quarto-journals/plos/blob/main/_extensions/plos/partials/before-body.tex

\maketitle

\begin{center}

\thispagestyle{fancy}

% Insert author names, affiliations and corresponding author email

\normalsize{Bjørn Peare
Bartholdy}\textsuperscript{1*}, \normalsize{Jørgen B.
Hasselstrøm}\textsuperscript{2}, \normalsize{Lambert K.
Sørensen}\textsuperscript{2}, \normalsize{Maia
Casna}\textsuperscript{1}, \normalsize{Menno
Hoogland}\textsuperscript{1}, \normalsize{Historisch Genootschap
Beemster}\textsuperscript{3}, \normalsize{Amanda G.
Henry}\textsuperscript{1}\vspace{12pt}

\rule{\textwidth}{1.8pt}\vspace{8pt}

\end{center}

\textsuperscript{\footnotesize{1}}\textit{\small{Department of
Archaeological Sciences, Leiden
University}}\\ \textsuperscript{\footnotesize{2}}\textit{\small{Department
of Forensic Medicine, Aarhus
University}}\\ \textsuperscript{\footnotesize{3}}\textit{\small{Historisch
Genootschap Beemster}}

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* \small{b.p.bartholdy@arch.leidenuniv.nl}

\vspace{12pt}

\begin{abstract}
Dental calculus is an excellent source of information on the dietary
patterns of past populations, including consumption of plant-based
items. The detection of plant-derived residues such as alkaloids and
their metabolites in dental calculus provides direct evidence of
consumption by individuals within a population. We conducted a study on
41 individuals from Middenbeemster, a 19th century rural Dutch
archaeological site. Skeletal and dental analysis was performed to
explore potential relationships between pathological conditions/lesions
and the presence of alkaloids. We also explored other factors
potentially affecting the detection of alkaloids, including sample
weight and skeletal preservation. Dental calculus was sampled and
analysed using ultra-high-performance liquid chromatography-tandem mass
spectrometry (UHPLC-ESI-MS/MS). We were able to detect nicotine,
cotinine, caffeine, theophylline, and salicylic acid. By detecting these
compounds we are able to show the consumption of tea and coffee and
smoking of tobacco on an individual scale, which is also confirmed by
historic documentation and identification of pipe notches in the
dentition. Nicotine and/or cotinine was present in 56\% of individuals
with at least one visible pipe notch. We find some influence of skeletal
preservation on the detection of alkaloids and salicylic acid, with
higher quantities of compounds extracted from well-preserved
individuals, and also observe a relationship between weight of the
calculus sample and raw quantity of the detected compounds, and we were
able to detect alkaloids in samples as small as 2 mg. We found
correlations between chronic maxillary sinusitis and the presence of
multiple alkaloids. We show that there are many limitations that will
need to be addressed going forward with this type of analysis, and
stress the need for more systematic research on the consumption of
alkaloid-containing items and their subsequent concentration and
preservation in dental calculus, in addition to how mode of consumption
may affect concentrations on different parts of the dentition. Despite
the limitations, this preliminary study illustrates the many benefits of
using calculus to target a variety of compounds that could have been
ingested as medicine or diet, or consumed in a different manner. This
method allows us to directly address specific individuals, which can be
especially useful in individuals that are not always well-documented in
historic documentation, such as rural populations, and especially
children and women.
\end{abstract}

\small \textbf{Keywords:}  dental calculus; LC-MS/MS; alkaloids; dental
pathology; sinusitis; caffeine; tobacco   \\

\vfill

\newgeometry{top=30mm,left=30mm}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Dental calculus has proven to be an excellent source of a wide variety
of information about our past. The increased accessibility and
advancement of methods in aDNA, paleoproteomics, and mass spectrometry,
has expanded our ability to identify biomarkers of diet and disease on
an increasingly large scale (Gismondi et al., 2020; Velsko et al., 2017;
Warinner et al., 2014).

One such collection of biomarkers is alkaloids, a plant-derived group of
compounds. Many alkaloids have important medicinal and psychoactive
effects in humans, and their direct detection, or detection of their
metabolites, is of great interest to archaeologists. Previous studies
have successfully recovered alkaloids in archaeological contexts,
including ceramics (Smith et al., 2018), pipes (Rafferty et al., 2012),
human hair (Echeverría \& Niemeyer, 2013; Ogalde et al., 2009), and even
dental calculus employing both targeted (Eerkens et al., 2018) and
untargeted approaches (Buckley et al., 2014; Gismondi et al., 2020).
Especially nicotine, the principal alkaloid in tobacco leaves, has been
widely studied in the archaeological record due to its apparent
stability and ability to survive over long periods of time (Eerkens et
al., 2018; Rafferty et al., 2012; Tushingham et al., 2013).

Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through ingestion of alkaloid-containing plants, whether
deliberate or accidental; and (2) passive diffusion as alkaloids and
other compounds are transferred from plasma to saliva, and then
gradually secreted into the oral cavity through the salivary glands in
the hours-to-days following ingestion (Cone \& Huestis, 2007). The
second pathway allows the identification of parent compounds that do not
enter the mouth (e.g.~injection), as long as they, or their metabolites,
are excreted through the saliva, thus eventually entering the oral
cavity.

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(Jin \& Yip, 2002; White, 1997). This may also allow various alkaloids
of dietary and medicinal origin to become incorporated in dental plaque.
Dental plaque undergoes frequent mineralisation events, ultimately
causing the entrapped alkaloids and their metabolites to become
preserved within the dental calculus. Barring intentional or accidental
removal of the calculus during life, burial, excavation, and
post-excavation cleaning, the alkaloids can then be detected by various
methods to show a record of consumption during life. Because drugs may
be transferred from plasma to saliva, there is often a close correlation
between drugs detected in oral fluid and blood, though there are
differences in detected concentrations (Cone \& Huestis, 2007; Milman et
al., 2011; Wille et al., 2009). This was also shown to be true for
dental calculus and blood (Sørensen et al., 2021), making dental
calculus a potentially useful substance for detecting ancient alkaloids
and other dietary compounds.

In this study we use a ultra-high-performance liquid
chromatography-tandem mass spectrometry (UHPLC-MS/MS) method that was
developed in a previous study on dental calculus from cadavers and
validated by comparing the results to compounds detected in the blood of
the same individuals (Sørensen et al., 2021). All compounds that were
detected in the blood were also detected in dental calculus, with
additional compounds present in dental calculus that were not present in
blood, suggesting that dental calculus represents a comprehensive
history of consumption over a long period of time (Sørensen et al.,
2021). We were able to detect both parent compounds and metabolites,
including caffeine, nicotine, theophylline, and cotinine, in the dental
calculus of individuals from a 19th century Dutch population from
Middenbeemster. By detecting these compounds we are able to show the
consumption of tea and coffee and smoking of tobacco on an individual
scale, which is also confirmed by historic documentation and
identification of pipe notches in the dentition.

\hypertarget{materials}{%
\section{Materials}\label{materials}}

The sample consists of 41 individuals from Middenbeemster, a 19th
century rural Dutch site. The village of Middenbeemster and the
surrounding Beemsterpolder was established in the beginning of the 17th
century, when the Beemster lake was drained to create more farmland,
mainly for the cultivation of cole seeds (de Vries 1978). In 1615, a
decision was made to build a church, and construction started in 1618
(Hakvoort 2013). The excavated cemetery is associated with the
Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
(Lemmers et al., 2013). Archival documents are available for those
buried between AD 1829 and 1866, when the majority of individuals were
interred. The main occupation of the inhabitants was dairy farming,
consisting largely of manual labour prior to the industrial revolution
(Aten et al., 2012; Palmer et al., 2016).

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour (Palmer et al., 2016), and there are
biological differences in dental calculus formation and drug metabolism
that are related to age and sex (Huang et al., 2023; Uno et al., 2017;
White, 1997). The sample consists of 27 males, 11 probable males, 2
probable females, and 1 female (Figure~\ref{fig-sample-demography}). We
selected males due to a higher occurrence of pipe notches and dental
calculus deposits than females (unpublished observation).

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-sample-demography-1.pdf}

@@

\end{figure}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{skeletal-analysis}{%
\subsection{Skeletal analysis}\label{skeletal-analysis}}

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits (Buikstra \& Ubelaker, 1994).
Age-at-death was estimated using dental wear, auricular and pubic
surface appearance, cranial suture closure, and epiphyseal fusion
(Brooks \& Suchey, 1990; Buckberry \& Chamberlain, 2002; Buikstra \&
Ubelaker, 1994; Lovejoy et al., 1985; Meindl \& Lovejoy, 1985), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

\hypertarget{paleopathology}{%
\subsubsection{Paleopathology}\label{paleopathology}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11.5pt,
  leqno]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[sfdefault]{roboto}
\usepackage{unicode-math}
\usepackage[a4paper,headheight=13pt,footskip=20pt,margin=2.54cm]{geometry}
\usepackage{indentfirst}

% Modify section headings
\usepackage{titlesec}
\titleformat{\section}{\normalfont\large\bfseries\centering}{}{.5em}{}
\titleformat{\subsection}{\normalfont\bfseries}{}{.5em}{}
\titleformat{\subsubsection}[runin]{\normalfont\itshape}{\thesubsubsection.}{.5em}{\addperiod}
\newcommand{\addperiod}[1]{#1.} %https://tex.stackexchange.com/questions/137193/adding-a-period-after-section

\setlength{\parindent}{1.5em}
\setlength{\parskip}{0pt plus 1pt}
\linespread{1.1}

% Adjust caption style
\usepackage[labelfont=bf,
    labelsep=endash,
    singlelinecheck=off,
    format=plain,
    margin=1.5cm,
    aboveskip=12pt,
    belowskip=12pt,
    font=footnotesize,
    justification=justified]{caption}

% Defining document colors
\usepackage{xcolor}
\definecolor{darkgray}{HTML}{808080}
\definecolor{mediumgray}{HTML}{6D6E70}
\definecolor{ligthgray}{HTML}{d9d9d9}
\definecolor{pciblue}{HTML}{346E92}
\definecolor{opengreen}{HTML}{77933c}

%\usepackage[colorlinks=true,linkcolor=pciblue,urlcolor=pciblue,citecolor=pciblue]{hyperref} % Not working - links in pdf not changing colour. 
                                                         %Probably because hyperref is no the last package to be loaded, which it needs to be.
%\urlstyle{same}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Multiproxy analysis exploring patterns of diet and disease in dental calculus and skeletal remains from a 19th century Dutch population},
  pdfauthor={Bjørn Peare Bartholdy; Jørgen B. Hasselstrøm; Lambert K. Sørensen; Maia Casna; Menno Hoogland; Historisch Genootschap Beemster; Amanda G. Henry},
  pdfkeywords={dental calculus; LC-MS/MS; alkaloids; dental pathology;
sinusitis; caffeine; tobacco},
  colorlinks=true,
  linkcolor={pciblue},
  filecolor={Maroon},
  citecolor={pciblue},
  urlcolor={pciblue},
  pdfcreator={LaTeX via pandoc}}

\title{}
\author{}
\date{}

\begin{document}

\setstretch{1.5}
\section{Introduction}\label{introduction}

Dental calculus has proven to be an excellent source of a wide variety
of information about our past. The increased accessibility and
advancement of methods in aDNA, paleoproteomics, and mass spectrometry,
has expanded our ability to identify biomarkers of diet and disease on
an increasingly large scale
(\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}; \citeproc{ref-velskoDentalCalculus2017}{Velsko et al., 2017};
\citeproc{ref-warinnerEvidenceMilk2014}{Warinner et al., 2014}).

One such collection of biomarkers is alkaloids, a plant-derived group of
compounds. Many alkaloids have important medicinal and psychoactive
effects in humans, and their direct detection, or detection of their
metabolites, is of great interest to archaeologists. Previous studies
have successfully recovered alkaloids in archaeological contexts,
including ceramics (\citeproc{ref-smithDetectionOpium2018}{Smith et al.,
2018}), pipes (\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et
al., 2012}), human hair
(\citeproc{ref-echeverriaNicotineHair2013}{Echeverría \& Niemeyer,
2013}; \citeproc{ref-ogaldeIdentificationPsychoactive2009}{Ogalde et
al., 2009}), and even dental calculus employing both targeted
(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018}) and
untargeted approaches (\citeproc{ref-buckleyDentalCalculus2014}{Buckley
et al., 2014};
\citeproc{ref-gismondiMultidisciplinaryApproach2020}{Gismondi et al.,
2020}). Especially nicotine, the principal alkaloid in tobacco leaves,
has been widely studied in the archaeological record due to its apparent
stability and ability to survive over long periods of time
(\citeproc{ref-eerkensDentalCalculus2018}{Eerkens et al., 2018};
\citeproc{ref-raffertyCurrentResearch2012}{Rafferty et al., 2012};
\citeproc{ref-tushinghamHuntergathererTobacco2013}{Tushingham et al.,
2013}).

Alkaloids may enter the oral cavity via two pathways: (1) direct
incorporation through ingestion of alkaloid-containing plants, whether
deliberate or accidental; and (2) passive diffusion as alkaloids and
other compounds are transferred from plasma to saliva, and then
gradually secreted into the oral cavity through the salivary glands in
the hours-to-days following ingestion
(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007}). The
second pathway allows the identification of parent compounds that do not
enter the mouth (e.g.~injection), as long as they, or their metabolites,
are excreted through the saliva, thus eventually entering the oral
cavity.

Many of the components involved in the formation and growth of dental
calculus originate from oral fluid. Proteins, bacteria, salts and other
compounds are transferred from saliva to biofilms on the tooth surface
(\citeproc{ref-jinSupragingivalCalculus2002}{Jin \& Yip, 2002};
\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). This may also
allow various alkaloids of dietary and medicinal origin to become
incorporated in dental plaque. Dental plaque undergoes frequent
mineralisation events, ultimately causing the entrapped alkaloids and
their metabolites to become preserved within the dental calculus.
Barring intentional or accidental removal of the calculus during life,
burial, excavation, and post-excavation cleaning, the alkaloids can then
be detected by various methods to show a record of consumption during
life. Because drugs may be transferred from plasma to saliva, there is
often a close correlation between drugs detected in oral fluid and
blood, though there are differences in detected concentrations
(\citeproc{ref-coneInterpretationOral2007}{Cone \& Huestis, 2007};
\citeproc{ref-milmanOralFluid2011}{Milman et al., 2011};
\citeproc{ref-willeRelationshipOral2009}{Wille et al., 2009}). This was
also shown to be true for dental calculus and blood
(\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}),
making dental calculus a potentially useful substance for detecting
ancient alkaloids and other dietary compounds.

In this study we use a ultra-high-performance liquid
chromatography-tandem mass spectrometry (UHPLC-MS/MS) method that was
developed in a previous study on dental calculus from cadavers and
validated by comparing the results to compounds detected in the blood of
the same individuals (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen
et al., 2021}). All compounds that were detected in the blood were also
detected in dental calculus, with additional compounds present in dental
calculus that were not present in blood, suggesting that dental calculus
represents a comprehensive history of consumption over a long period of
time (\citeproc{ref-sorensenDrugsCalculus2021}{Sørensen et al., 2021}).
We were able to detect both parent compounds and metabolites, including
caffeine, nicotine, theophylline, and cotinine, in the dental calculus
of individuals from a 19th century Dutch population from Middenbeemster.
By detecting these compounds we are able to show the consumption of tea
and coffee and smoking of tobacco on an individual scale, which is also
confirmed by historic documentation and identification of pipe notches
in the dentition.

\section{Materials}\label{materials}

The sample consists of 41 individuals from Middenbeemster, a 19th
century rural Dutch site. The village of Middenbeemster and the
surrounding Beemsterpolder was established in the beginning of the 17th
century, when the Beemster lake was drained to create more farmland,
mainly for the cultivation of cole seeds (de Vries 1978). In 1615, a
decision was made to build a church, and construction started in 1618
(Hakvoort 2013). The excavated cemetery is associated with the
Keyserkerk church, where the inhabitants of the Middenbeemster village
and the surrounding Beemsterpolder were buried between AD 1615 and 1866
(\citeproc{ref-lemmersMiddenbeemster2013}{Lemmers et al., 2013}).
Archival documents are available for those buried between AD 1829 and
1866, when the majority of individuals were interred. The main
occupation of the inhabitants was dairy farming, consisting largely of
manual labour prior to the industrial revolution
(\citeproc{ref-aten400Jaar2012}{Aten et al., 2012};
\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}).

For our sample, we preferentially selected males from the middle adult
age category (35-49 years) to minimise the effect of confounding
cultural and biological factors. Previous research on Middenbeemster has
shown a gendered division of labour
(\citeproc{ref-palmerActivityReconstruction2016}{Palmer et al., 2016}),
and there are biological differences in dental calculus formation and
drug metabolism that are related to age and sex
(\citeproc{ref-huangDecipheringGenetic2023}{Huang et al., 2023};
\citeproc{ref-unoSexAgedependent2017}{Uno et al., 2017};
\citeproc{ref-whiteDentalCalculus1997}{White, 1997}). The sample
consists of 27 males, 11 probable males, 2 probable females, and 1
female (Figure~\ref{fig-sample-demography}). We selected males due to a
higher occurrence of pipe notches and dental calculus deposits than
females (unpublished observation).

@@

\centering{

\includegraphics{../figures/fig-sample-demography-1.pdf}

@@

\end{figure}%

\section{Methods}\label{methods}

\subsection{Skeletal analysis}\label{skeletal-analysis}

Demographic and pathological analyses were conducted in the Laboratory
for Human Osteoarchaeology at Leiden University. Sex was estimated using
cranial and pelvic morphological traits
(\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994}). Age-at-death
was estimated using dental wear, auricular and pubic surface appearance,
cranial suture closure, and epiphyseal fusion
(\citeproc{ref-SucheyBrooks1990}{Brooks \& Suchey, 1990};
\citeproc{ref-buckberryAuricular2002}{Buckberry \& Chamberlain, 2002};
\citeproc{ref-Standards1994}{Buikstra \& Ubelaker, 1994};
\citeproc{ref-lovejoyAuricular1985}{Lovejoy et al., 1985};
\citeproc{ref-meindlSutureClosure1985}{Meindl \& Lovejoy, 1985}), and
divided into the following categories: early young adult (18-24 years),
late young adult (25-34 years), middle adult ( 35-49 years), old adult
(50+ years). Preservation was visually scored on a four-stage scale
(excellent, good, fair, poor) based on the surface condition of the
bones and the extent of taphonomic degradation.

\subsubsection{Paleopathology}\label{paleopathology}
